---
title: "Simulation Study: VAM Model Recovery"
subtitle: "Replicating and extending Kim (2022) Section 4-5"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(cmdstanr)
library(posterior)
library(MASS)
```

## Overview

This simulation study mirrors Kim (2022) Section 4 to evaluate how well each
model recovers true teacher quality under three data-generating scenarios:

- **Scenario A** — Student parameters are *invariant*: each student has their
  own stable alpha_i and a_i across grades. Favours SWVAM.
- **Scenario B** — Student parameters are *identical*: all students share the
  same parameters. Favours CVA (traditional VAM).
- **Scenario C** — Student parameters are *random*: new parameters drawn
  independently each grade. Neither model's assumptions hold.

Performance is measured by (1) correlation of estimates with true teacher
quality and (2) proportion of teachers correctly assigned to quality quartile,
following Kim (2022) Tables 2 and 3.

---

## 1. Data-generating process

```{r dgp}
simulate_data <- function(
    n_teachers = 20,
    n_students = 200,
    n_grades   = 3,
    n_schools  = 4,
    scenario   = "A",
    rho        = 0.7,
    seed       = 141229
) {
  set.seed(seed)

  # ── True effects ────────────────────────────────────────────────────────────
  theta_true     <- rnorm(n_teachers)
  school_effects <- rnorm(n_schools, 0, 0.3)

  # ── School assignments ──────────────────────────────────────────────────────
  # Teachers evenly distributed across schools
  teacher_school <- rep(1:n_schools, length.out = n_teachers)

  # Students evenly distributed across schools
  student_school <- rep(1:n_schools, length.out = n_students)

  # ── Student-teacher assignments (within school) ─────────────────────────────
  assign_one_grade <- function() {
    assignment <- integer(n_students)
    for (s in 1:n_schools) {
      school_students <- which(student_school == s)
      school_teachers <- which(teacher_school == s)
      assignment[school_students] <- sample(
        rep(school_teachers, length.out = length(school_students))
      )
    }
    assignment
  }

  teacher_assign <- matrix(
    replicate(n_grades, assign_one_grade()),
    nrow = n_students, ncol = n_grades
  )

  # ── Student parameters ──────────────────────────────────────────────────────
  if (scenario == "A") {
    # Each student has own params, stable across grades
    alpha_i   <- rlnorm(n_students, log(0.2), log(2) / 2)
    a_i       <- rlnorm(n_students, log(1.0), log(2) / 2)
    alpha_mat <- matrix(rep(alpha_i, n_grades), nrow = n_students)
    a_mat     <- matrix(rep(a_i,     n_grades), nrow = n_students)

  } else if (scenario == "B") {
    # All students share one set of parameters
    alpha_shared <- rlnorm(1, log(0.2), log(2) / 2)
    a_shared     <- rlnorm(1, log(1.0), log(2) / 2)
    alpha_mat    <- matrix(alpha_shared, nrow = n_students, ncol = n_grades)
    a_mat        <- matrix(a_shared,     nrow = n_students, ncol = n_grades)

  } else if (scenario == "C") {
    # New params drawn for every student-grade combination
    alpha_mat <- matrix(
      rlnorm(n_students * n_grades, log(0.2), log(2) / 2),
      nrow = n_students
    )
    a_mat <- matrix(
      rlnorm(n_students * n_grades, log(1.0), log(2) / 2),
      nrow = n_students
    )
  }

  # ── AR(1) cross-wave covariance structure ───────────────────────────────────
  # Captures persistent student ability across grades — gives TVAAS
  # something to learn that CVA/SWVAM cannot exploit
  Sigma_base <- rho^abs(outer(1:n_grades, 1:n_grades, "-"))

  # ── Generate outcomes ───────────────────────────────────────────────────────
  rows <- vector("list", n_students * n_grades)
  k    <- 1L

  for (i in seq_len(n_students)) {
    j_vec <- teacher_assign[i, ]

    # School effect from first grade teacher (stable within school)
    s <- teacher_school[j_vec[1]]

    # Mean vector across grades — teacher + school
    mu_vec <- sapply(seq_len(n_grades), function(g) {
      alpha_mat[i, g] * theta_true[j_vec[g]] + school_effects[s]
    })

    # Student-specific precision scales covariance
    # Use first grade a_i as the baseline precision
    sigma_i <- 1.0 / sqrt(a_mat[i, 1])
    Sigma_i <- Sigma_base * sigma_i^2

    # Draw all grades jointly — this is the key change for TVAAS
    scores_i <- MASS::mvrnorm(1, mu_vec, Sigma_i)

    for (g in seq_len(n_grades)) {
      rows[[k]] <- list(
        student_id = i,
        teacher_id = j_vec[g],
        school_id  = s,
        grade      = g,
        y          = scores_i[g]
      )
      k <- k + 1L
    }
  }

  data <- bind_rows(rows)

  list(
    data           = data,
    theta_true     = theta_true,
    school_effects = school_effects,
    teacher_school = teacher_school,
    student_school = student_school
  )
}
```

---

## 2. Stan data builders

```{r stan_builders}
# CVA stan_data — no real covariates in simulation (Kim Section 4 footnote),
# so we pass a dummy zero column for X and zero vector for y_prev
make_sim_cva_data <- function(df, n_teachers) {
  list(
    N          = nrow(df),
    J          = n_teachers,
    K          = 1L,
    y          = df$y,
    y_prev     = rep(0.0, nrow(df)),
    X          = matrix(0.0, nrow(df), 1L),
    teacher_id = df$teacher_id
  )
}

# SWVAM stan_data
make_sim_swvam_data <- function(df, n_teachers, n_students) {
  list(
    N          = nrow(df),
    I          = n_students,
    J          = n_teachers,
    K          = 1L,
    TT         = max(df$grade),
    y          = df$y,
    y_prev     = rep(0.0, nrow(df)),
    X          = matrix(0.0, nrow(df), 1L),
    student_id = df$student_id,
    teacher_id = df$teacher_id,
    time_id    = df$grade
  )
}

make_sim_composite_data <- function(df, n_teachers, n_students, teacher_school) {

  # Peer achievement — mean score per school-grade
  df <- df |>
    group_by(school_id, grade) |>
    mutate(P = mean(y)) |>
    ungroup()

  # School covariate — school membership as single column
  S <- matrix(as.numeric(teacher_school), ncol = 1)

  list(
    N          = nrow(df),
    I          = n_students,
    J          = n_teachers,
    K          = 1L,
    L          = 1L,
    TT         = max(df$grade),
    y          = df$y,
    y_prev     = rep(0.0, nrow(df)),
    P          = df$P,
    X          = matrix(0.0, nrow(df), 1L),
    S          = S,
    student_id = df$student_id,
    teacher_id = df$teacher_id,
    time_id    = df$grade
  )
}

# ── Add to data builders chunk ─────────────────────────────────────────────
make_sim_tvaas_data <- function(df, n_teachers, n_students, n_grades) {

  Y_wide <- df |>
    arrange(student_id, grade) |>
    pivot_wider(
      id_cols      = student_id,
      names_from   = grade,
      values_from  = y,
      names_prefix = "grade_"
    ) |>
    arrange(student_id) |>
    dplyr::select(-student_id) |>
    as.matrix()

  T_wide <- df |>
    arrange(student_id, grade) |>
    pivot_wider(
      id_cols      = student_id,
      names_from   = grade,
      values_from  = teacher_id,
      names_prefix = "grade_"
    ) |>
    arrange(student_id) |>
    dplyr::select(-student_id) |>
    as.matrix()

  obs_mask  <- matrix(1L, nrow = n_students, ncol = n_grades)
  n_obs_vec <- rep(n_grades, n_students)

  list(
    N          = n_students,
    J          = n_teachers,
    G          = n_grades,
    Y          = Y_wide,
    obs        = obs_mask,
    n_obs      = n_obs_vec,
    teacher_id = T_wide
  )
}
```

---

## 3. Recovery metrics

```{r metrics}
evaluate_recovery <- function(theta_estimated, theta_true) {
  # Correlation with true teacher quality
  r <- cor(theta_estimated, theta_true, use = "complete.obs")

  # Proportion correctly assigned to quality quartile (Kim Table 2)
  q_breaks_true <- quantile(theta_true,      probs = 0:4 / 4)
  q_breaks_est  <- quantile(theta_estimated, probs = 0:4 / 4)

  true_q <- cut(theta_true,      breaks = q_breaks_true,
                labels = 1:4, include.lowest = TRUE)
  est_q  <- cut(theta_estimated, breaks = q_breaks_est,
                labels = 1:4, include.lowest = TRUE)

  pct_correct <- mean(true_q == est_q, na.rm = TRUE)

  list(correlation = round(r, 4), pct_correct_quartile = round(pct_correct, 4))
}
```

---

## 4. Load compiled models

```{r load_models}
# Compile once — reuses cached executables on subsequent runs
mod_cva   <- cmdstan_model("cva_model.stan")
mod_swvam <- cmdstan_model("swvam_model.stan")
mod_composite <- cmdstan_model("composite_model.stan")
mod_tvaas <- cmdstan_model("tvaas_model.stan")
```

---

## 5. Single-trial runner

```{r single_trial}
run_trial <- function(scenario, n_grades, n_teachers = 20,
                      n_students = 200, n_schools = 4, 
                      rho = 0.7, seed = 1L) {

  sim <- simulate_data(
    n_teachers = n_teachers,
    n_students = n_students,
    n_grades   = n_grades,
    n_schools  = n_schools,
    scenario   = scenario,
    rho        = rho,
    seed       = seed
  )

  # ── Fit CVA ───────────────────────────────────────────────────────────────
  fit_cva <- mod_cva$sample(
    data          = make_sim_cva_data(sim$data, n_teachers),
    seed          = seed, chains = 1,
    iter_warmup   = 500,
    iter_sampling = 1000,
    refresh = 0, show_messages = FALSE
  )
  theta_cva <- fit_cva$summary("theta")$mean
  if (sd(theta_cva) < 1e-6) return(NULL)

  # ── Fit SWVAM ─────────────────────────────────────────────────────────────
  fit_swvam <- mod_swvam$sample(
    data          = make_sim_swvam_data(sim$data, n_teachers, n_students),
    seed          = seed, chains = 1,
    iter_warmup   = 500,
    iter_sampling = 1000,
    refresh = 0, show_messages = FALSE
  )
  theta_swvam <- fit_swvam$summary("theta")$mean
  if (sd(theta_swvam) < 1e-6) return(NULL)

  # ── Fit Composite ─────────────────────────────────────────────────────────
  fit_comp <- mod_composite$sample(
  data = make_sim_composite_data(sim$data, n_teachers, 
                                 n_students, sim$teacher_school),
    seed          = seed, chains = 1,
    iter_warmup   = 500,
    iter_sampling = 1000,
    refresh = 0, show_messages = FALSE
  )
  theta_comp <- fit_comp$summary("theta")$mean
  if (sd(theta_comp) < 1e-6) return(NULL)
    # ── Fit TVAAS ─────────────────────────────────────────────────────────────
  fit_tvaas <- mod_tvaas$sample(
    data          = make_sim_tvaas_data(sim$data, n_teachers,
                                        n_students, n_grades),
    seed          = seed, chains = 1,
    iter_warmup   = 500,
    iter_sampling = 1000,
    refresh       = 0, show_messages = FALSE
  )

  if (any(fit_tvaas$diagnostic_summary()$ebfmi < 0.3)) return(NULL)

  theta_tvaas <- fit_tvaas$summary("theta")$mean
  if (sd(theta_tvaas) < 1e-6) return(NULL)

  # ── Save theta summaries ───────────────────────────────────────────────────
  extract_theta <- function(fit) {
    fit$summary("theta", mean, sd, ~quantile(.x, c(0.05, 0.95))) |>
      rename(mean = mean, sd = sd, q05 = `5%`, q95 = `95%`)
  }

  saveRDS(
    list(
      scenario   = scenario,
      n_grades   = n_grades,
      seed       = seed,
      theta_true = sim$theta_true,
      cva        = extract_theta(fit_cva),
      swvam      = extract_theta(fit_swvam),
      composite  = extract_theta(fit_comp),
      tvaas      = extract_theta(fit_tvaas)
    ),
    sprintf("stan_fits/thetas_%s_g%d_seed%d.rds", scenario, n_grades, seed)
  )

  # ── Evaluate ──────────────────────────────────────────────────────────────
  rec_cva   <- evaluate_recovery(theta_cva,   sim$theta_true)
  rec_swvam <- evaluate_recovery(theta_swvam, sim$theta_true)
  rec_comp  <- evaluate_recovery(theta_comp,  sim$theta_true)
  rec_tvaas <- evaluate_recovery(theta_tvaas, sim$theta_true)
  
    tibble(
    scenario  = scenario,
    n_grades  = n_grades,
    trial     = seed,
    cor_cva   = rec_cva$correlation,
    cor_swvam = rec_swvam$correlation,
    cor_comp  = rec_comp$correlation,
    cor_tvaas = rec_tvaas$correlation,
    pct_cva   = rec_cva$pct_correct_quartile,
    pct_swvam = rec_swvam$pct_correct_quartile,
    pct_comp  = rec_comp$pct_correct_quartile,
    pct_tvaas = rec_tvaas$pct_correct_quartile,
    swvam_minus_cva_cor  = rec_swvam$correlation          - rec_cva$correlation,
    comp_minus_cva_cor   = rec_comp$correlation           - rec_cva$correlation,
    tvaas_minus_cva_cor  = rec_tvaas$correlation          - rec_cva$correlation,
    swvam_minus_cva_pct  = rec_swvam$pct_correct_quartile - rec_cva$pct_correct_quartile,
    comp_minus_cva_pct   = rec_comp$pct_correct_quartile  - rec_cva$pct_correct_quartile,
    tvaas_minus_cva_pct  = rec_tvaas$pct_correct_quartile - rec_cva$pct_correct_quartile
  )
}
```

---

## 6. Full simulation study

Runs 25 trials per scenario × grade combination (reduced from Kim's 100 for
computational feasibility; increase `n_trials` for final results).

```{r run_simulation, cache=TRUE}
N_TRIALS  <- 100
SCENARIOS <- c("A", "C")  
GRADES    <- c(2L, 4L, 6L)

all_results <- list()

for (sc in SCENARIOS) {
  for (ng in GRADES) {
    message(sprintf("Running scenario %s, %d grades...", sc, ng))

    trial_results <- vector("list", N_TRIALS)
    valid_trials  <- 0L
    seed          <- 1L

    while (valid_trials < N_TRIALS) {
      result <- tryCatch(
        run_trial(scenario = sc, n_grades = ng, seed = seed),
        error = function(e) NULL
      )
      if (!is.null(result)) {
        valid_trials <- valid_trials + 1L
        trial_results[[valid_trials]] <- result
      }
      seed <- seed + 1L
    }

    all_results[[paste(sc, ng)]] <- bind_rows(trial_results)
    
    # Save after each scenario-grade combination completes
    results_df <- bind_rows(all_results)
    saveRDS(results_df, "simulation_results.rds")
    message(sprintf("  Saved. Seeds used: %d", seed - 1L))
  }
}

results_df <- bind_rows(all_results)
saveRDS(results_df, "simulation_results.rds")
```


```{r}
results_df <- readRDS("simulation_results.rds")
results_df |> count(scenario, n_grades)
```

---

## 7. Results tables (replicating Kim Tables 2 & 3)

```{r results_tables}
results_df |>
  pivot_longer(
    cols = c(swvam_minus_cva_cor, comp_minus_cva_cor, tvaas_minus_cva_cor,
         swvam_minus_cva_pct, comp_minus_cva_pct, tvaas_minus_cva_pct),
    names_to  = "series",
    values_to = "advantage"
  ) |>
  mutate(
    metric   = if_else(str_detect(series, "cor"),
                       "Correlation with true θ",
                       "Proportion correct quartile"),
    model = case_when(
  str_detect(series, "swvam") ~ "SWVAM",
  str_detect(series, "comp")  ~ "Composite",
  str_detect(series, "tvaas") ~ "TVAAS"
),
    scenario = recode(scenario,
                      A = "A: Invariant",
                      B = "B: Identical",
                      C = "C: Random")
  ) |>
  ggplot(aes(x = factor(n_grades), y = advantage, fill = model)) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = "grey50", linewidth = 0.4) +
  geom_boxplot(
    alpha         = 0.7,
    outlier.size  = 0.8,
    outlier.alpha = 0.5,
    position      = position_dodge(width = 0.75),
    linewidth     = 0.4
  ) +
  facet_grid(scenario ~ metric) +
  scale_fill_manual(values = c(
  "SWVAM"     = "#1D9E75",
  "Composite" = "#D85A30",
  "TVAAS"     = "#7F77DD"
)) +
  labs(
    title    = "Model advantage over CVA across simulation scenarios",
    subtitle = "Positive = model better than CVA; negative = CVA better",
    x        = "Grades of observation per student",
    y        = "Model minus CVA",
    fill     = "Model"
  ) +
  theme_minimal(base_size = 11) +
 # Add these to your theme() call to polish it up
theme(
  panel.grid.minor  = element_blank(),
  strip.text.x      = element_text(face = "bold"),
  strip.text.y      = element_text(face = "bold", angle = 0),  # readable row labels
  legend.position   = "bottom",                                 # move legend below
  panel.spacing     = unit(0.8, "lines")                        # more breathing room
)
    
```
```{r}
results_df |> 
  group_by(scenario, n_grades) |>
  summarise(mean(comp_minus_cva_cor), sd(comp_minus_cva_cor))
```
```{r}
# How often does composite actually beat CVA?
results_df |>
  group_by(scenario, n_grades) |>
  summarise(
    pct_comp_beats_cva = mean(comp_minus_cva_cor > 0),
    pct_tvaas_beats_cva = mean(tvaas_minus_cva_cor > 0)
  )
```

---

## 8. Visualisation

```{r plot_results, fig.width=10, fig.height=5}
results_df |>
  pivot_longer(
    cols      = c(swvam_minus_cva_cor, swvam_minus_cva_pct),
    names_to  = "metric",
    values_to = "swvam_advantage"
  ) |>
  mutate(
    metric   = recode(metric,
                      swvam_minus_cva_cor = "Correlation with true θ",
                      swvam_minus_cva_pct = "Proportion correct quartile"),
    scenario = recode(scenario,
                      A = "A: Invariant\n(favours SWVAM)",
                      B = "B: Identical\n(favours CVA)",
                      C = "C: Random\n(neither)")
  ) |>
  ggplot(aes(x = factor(n_grades), y = swvam_advantage, fill = scenario)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8) +
  facet_wrap(~ metric) +
  scale_fill_manual(values = c(
    "A: Invariant\n(favours SWVAM)" = "#1D9E75",
    "B: Identical\n(favours CVA)"   = "#378ADD",
    "C: Random\n(neither)"          = "#7F77DD"
  )) +
  labs(
    title    = "SWVAM advantage over CVA across simulation scenarios",
    subtitle = "Positive values = SWVAM better; negative = CVA better",
    x        = "Grades of observation per student",
    y        = "SWVAM minus CVA",
    fill     = "Scenario"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

```{r}
RHO_VALUES <- c(0.3, 0.5, 0.7, 0.9)
N_TRIALS_RHO <- 25
SCENARIOS_RHO <- c("A", "C")
GRADES_RHO <- c(2L, 4L, 6L)

rho_results <- list()

for (rho_val in RHO_VALUES) {
  for (sc in SCENARIOS_RHO) {
    for (ng in GRADES_RHO) {
      message(sprintf("rho=%.1f, scenario=%s, grades=%d", rho_val, sc, ng))
      
      trial_results <- vector("list", N_TRIALS_RHO)
      valid_trials <- 0L
      seed <- 1L
      
      while (valid_trials < N_TRIALS_RHO) {
        result <- tryCatch(
          run_trial(scenario = sc, n_grades = ng,
                    n_teachers = 20, n_students = 200,
                    n_schools = 4, rho = rho_val, seed = seed),
          error = function(e) NULL
        )
        if (!is.null(result)) {
          result$rho <- rho_val
          valid_trials <- valid_trials + 1L
          trial_results[[valid_trials]] <- result
        }
        seed <- seed + 1L
      }
      
      rho_results[[paste(rho_val, sc, ng)]] <- bind_rows(trial_results)
      rho_df <- bind_rows(rho_results)
      saveRDS(rho_df, "rho_sensitivity_results.rds")
    }
  }
}
```

```{r}
rho_df <- readRDS("rho_sensitivity_results.rds")
rho_df |> count(rho, scenario, n_grades)
```




```{r}
rho_df |>
  group_by(rho, scenario, n_grades) |>
  summarise(
    mean_tvaas = mean(tvaas_minus_cva_cor),
    mean_comp  = mean(comp_minus_cva_cor),
    .groups = "drop"
  ) |>
  pivot_longer(cols = c(mean_tvaas, mean_comp),
               names_to = "model", values_to = "advantage") |>
  mutate(
    model    = recode(model, mean_tvaas = "TVAAS", mean_comp = "Composite"),
    scenario = recode(scenario, A = "A: Invariant", C = "C: Random")
  ) |>
  ggplot(aes(x = rho, y = advantage, color = model)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  facet_grid(scenario ~ n_grades) +
  labs(
    title    = "Model advantage over CVA by AR(1) correlation strength",
    subtitle = "Positive = model better than CVA",
    x        = "Rho (AR(1) correlation)",
    y        = "Model minus CVA correlation with true θ"
  ) +
  theme_minimal(base_size = 11)
```

```{r}
rho_df |>
  group_by(rho, scenario, n_grades) |>
  summarise(
    mean_tvaas = mean(tvaas_minus_cva_cor),
    se_tvaas   = sd(tvaas_minus_cva_cor) / sqrt(n()),
    .groups = "drop"
  ) |>
  mutate(scenario = recode(scenario, A = "A: Invariant", C = "C: Random")) |>
  ggplot(aes(x = rho, y = mean_tvaas, color = factor(n_grades))) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_ribbon(aes(ymin = mean_tvaas - se_tvaas,
                  ymax = mean_tvaas + se_tvaas,
                  fill = factor(n_grades)), alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  geom_vline(xintercept = 0.7, linetype = "dotted", 
             colour = "grey30", linewidth = 0.5) +
  annotate("text", x = 0.72, y = -0.01, 
           label = "ECLS-K rho ≈ 0.7", hjust = 0, size = 3) +
  facet_wrap(~ scenario) +
  scale_color_manual(values = c("2" = "#378ADD", "4" = "#1D9E75", "6" = "#D85A30")) +
  scale_fill_manual(values  = c("2" = "#378ADD", "4" = "#1D9E75", "6" = "#D85A30")) +
  labs(
    title    = "TVAAS advantage over CVA scales with AR(1) correlation",
    subtitle = "Shaded regions = ±1 SE across 25 trials",
    x        = "Rho (AR(1) correlation)",
    y        = "TVAAS minus CVA (correlation with true θ)",
    color    = "Grades",
    fill     = "Grades"
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        legend.position  = "bottom")
```




